---
title: "Credit Risk Analysis using Regression"
author: "Gabriel ISSA"
date: "2026-10-07"
output: html_document
---

Package set up  
```{r setup, message=FALSE, warning=FALSE}
#install.package("stringr")
#install.package("stargazer")
#install.packages("ggplot2")
#install.packages("dyplr")
#install.packages("tidyr")
#install.packages("corrplot")
#library(stargazer)
library(stringr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrplot)
```


# Probability models conception

## Model design

The first step we need to do is to import the data and decide which variables we want to include in the model
```{r}
raw_df = read.csv("raw_data/credit-1.csv", stringsAsFactors = TRUE)
summary(raw_df)
```
#### Dependent variable : 

- *default* : 1 if the client made a default 0 else 

#### Raw variables to include : 

- *senority* : (negative) Longer job tenure signals stable income and lower risk of sudden job loss

- *age* : (negative) older borrowers are more experienced when managing their money and often have more stable financial situation which lead to lower default risk
  
#### Variables to transform : 

- *Is_house_owner* is equal to 1 if the client own a house 
  - Intuition : (negative) Homeowners have a valuable asset and may have already loan money to buy it so I expect lower default.

- *is_married* is equal to 1 if the client is married else 0 
  - Intuition : (negative) married couple have to income to reimburse a debt and a more stable situation which imply lower default risk. 
    
- *has_fixed_job* : 1 if job = "fixed" else 0 
  - Intuition : (negative) a full time salary job mean stable income and so less financial situation volatility

- *is_freelance* : 1 if job = "freelance" else 0 
  - Intuition : (positive) a freelance job is synonym of volatile income and changing financial situation which induce higher default risk 
    
- *is_parttime* : part-time job are often mean limited and unstable income 
- *reimbursement_cap* : `expenses / income`
  - Intution : (positive) High expenses leave little disposable income to service the loan, so the expenses-to-income ratio should be positively related to default.

- *income_to_total_debt* : `income / (debt + amount)`
  - Intuition : (negative) the higher is the debt compared to income the hardest it will be for the client to reimburse it 

- *savings_to_total_debt* : `savings / (debt + amount)`
  - Intuition : (negative) Savings are a liquidity cushion that lets a borrower keep paying during a temporary income shock, so the lower it is compared to debt the higher the higher the risk

- `log(1+income)` : 
  - Intuition : (negative) Higher income means more capacity to absorb repayments and unexpected shocks, so I expect a negative sign. I use the log because the effect shows diminishing returns.

- `log(1+asset)` : 
  - Intuition : (negative) same principle than income

#### To exclude of the analysis : 

- *id* : identifier of the client


### Final regression equation : 
This corresponds to the following final equation : 
$$
\begin{aligned}
\text{default}_i =\;& \beta_0
+ \beta_1 \text{seniority}_i
+ \beta_2 \text{age}_i \\
&+ \beta_3 \text{Is_house_owner}_i
+ \beta_4 \text{Is_married}_i \\
&+ \beta_5 \text{has_fixed_job}_i
+ \beta_6 \text{is_freelance}_i
+ \beta_7 \text{is_parttime}_i \\
&+ \beta_8 \text{reimbursement_cap}_i
+ \beta_9 \text{income_to_total_debt}_i \\
&+ \beta_{10} \text{savings_to_total_debt}_i \\
&+ \beta_{11} \ln(1+\text{income}_i)
+ \beta_{12} \ln(1+\text{assets}_i)
+ \varepsilon_i
\end{aligned}
$$


## Data Preparation 

The data needs to be cleaned and transformed to correspond to the designed model
```{r}
df = read.csv("raw_data/credit-1.csv", stringsAsFactors = TRUE)
df$Is_house_owner <- as.numeric(df$home == "owner")
df$Is_married <- as.numeric(df$marital == "married")
df$has_fixed_job <- as.numeric(df$job == "fixed")
df$is_freelance <- as.numeric(df$job == "freelance")
df$is_parttime <- as.numeric(df$job == "partime")
df$reimbursment_cap <- df$expenses/df$income
df$income_to_total_debt <- df$income/(df$debt+df$amount)
df$savings_to_total_debt <- df$savings/(df$debt+df$amount)
df$log_income <- log(1 + df$income)
df$log_assets <- log(1 + df$assets)

df = df[,c("default","seniority","age","Is_house_owner","Is_married","has_fixed_job","is_freelance","is_parttime","reimbursment_cap","income_to_total_debt","savings_to_total_debt","log_income","log_assets")]
```


Check colinearity and data :

The colinearity between all the variables needs to be computed to check the multicolinearity hypothesis and indicate if the dependant variable has a linear relationship with the independent variables. 
```{r}
vars <- c("default", "seniority", "age", "Is_house_owner","Is_married","has_fixed_job", "is_freelance", "is_parttime","reimbursment_cap", "income_to_total_debt","savings_to_total_debt","log_income", "log_assets")

cor_mat <- round(cor(df[, vars], use = "complete.obs"), 2)

corrplot(cor_mat,
         method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.55,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix", mar = c(0, 0, 2, 0))

```

# Model creation

## LPM model estimating

```{r}
LPM <- lm(default ~ seniority + age + Is_house_owner + Is_married + has_fixed_job + is_freelance + is_parttime + reimbursment_cap + income_to_total_debt + savings_to_total_debt + log_income + log_assets, df) 
summary(LPM)
```

## Logit model estimating 

```{r}
logit <- glm(default ~ seniority + age + Is_house_owner + Is_married + has_fixed_job + is_freelance + is_parttime + reimbursment_cap + income_to_total_debt + savings_to_total_debt + log_income + log_assets, df, family = binomial(link = "logit")) 
summary(logit)
```

## Probit estimating 

```{r}
probit <- glm(default ~ seniority + age + Is_house_owner + Is_married + has_fixed_job + is_freelance + is_parttime + reimbursment_cap + income_to_total_debt + savings_to_total_debt + log_income + log_assets, df, family = binomial( link = "probit"))
summary(probit)
```

## Interpretation :

- *seniority*  : negative coefficient as expected, significant at 99% level
- *age* : positive and significant at 99% level, meaning that the older is the client the higher is the higher is the default risk. May be due to bigger financial pressure due to children for age range 35 - 60 yo and less financial capacities for age range > 60 yo
- *Is_house_owner* : negative as expected, significant at 95% level of confidence 
- *Is_married* : not significant, may be due to high financial pressure for families 
- *has_fixed_job* : negative as expected, significant at 95% level of confidence
- *is_freelance* : not significant, being freelance may induce more volatility but is often related to higher incomes and more flexibility 
- *is_parttime* : positive as expected, significant at 99% level of confidence 
*reimbursment_cap* : not significant, may be due to the fact that expenses changes when people loan money, so the reimbursment_cap ratio is not relevant anymore 
- *income_to_total_debt* : negative and large meaning it has a big impact on the default capacity, which make sense because the risk of default is related to the value of the debt compared to the ability to reimburse this debt 
- *savings_to_total_debt* : surprisingly positive, may be due to colinearity with income_to_total_debt 
- *log_income* and *log_asset* : both negative and significant at 95% level of confidence

# Application of the model to 10 personas  

We will know use the 3 models created to estimate the probabilities of default of 10 clients that asked for a loan. 

## Data Preparation
```{r}
to_predict <- read.csv("raw_data/prospects.csv",stringsAsFactors = TRUE)
summary(to_predict)

to_predict$Is_house_owner <- as.numeric(to_predict$home == "owner")
to_predict$Is_married <- as.numeric(to_predict$marital == "married")
to_predict$has_fixed_job <- as.numeric(to_predict$job == "fixed")
to_predict$is_freelance <- as.numeric(to_predict$job == "freelance")
to_predict$is_parttime <- as.numeric(to_predict$job == "partime")
to_predict$reimbursment_cap <- to_predict$expenses/to_predict$income
to_predict$income_to_total_debt <- to_predict$income/(to_predict$debt+to_predict$amount)
to_predict$savings_to_total_debt <- to_predict$savings/(to_predict$debt+to_predict$amount)
to_predict$log_income <- log(1 + to_predict$income)
to_predict$log_assets <- log(1 + to_predict$assets)

to_predict = to_predict[,c("seniority","age","Is_house_owner","Is_married","has_fixed_job","is_freelance","is_parttime","reimbursment_cap","income_to_total_debt","savings_to_total_debt","log_income","log_assets","amount","prospect")]

```

Predict the probabilities for the prospect using the three different models
```{r}
to_predict$p_LPM <- pmin(1,pmax(0, predict(LPM, newdata = to_predict))) 
to_predict$p_logit <- predict(logit, newdata = to_predict, type = "response" )
to_predict$p_probit <- predict(probit, newdata = to_predict, type = "response")
```

## Models estimated probabilities comparison :
```{r}
probabilities <- to_predict %>%
  select(prospect, p_LPM, p_logit, p_probit) %>%
  pivot_longer(
    cols = c(p_LPM, p_logit, p_probit),
    names_to = "Model",
    values_to = "Probability"
  )

probabilities_comparison_plot <- ggplot(probabilities,
       aes(x = prospect, y = Probability, fill = Model)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Predicted Default Probability by Prospect",
    x = "Prospect",
    y = "Predicted probability",
    fill = "Model"
  ) +
  theme_minimal()

print(probabilities_comparison_plot)

ggsave( filename = "Output_plots/probabilities_comparison.png", plot = probabilities_comparison_plot , width = 9, height = 6, dpi = 300 )
```

# Application of the result to a 1- year loan 10% IR

We will know consider they asked for a 1-year loan with 10% interest rate, considering no cost of borrowing and no other variable costs : 

Earning without default : 
$$
\begin{aligned} Earning \ without \ default = \;& 0.10 *  Loan \ amount
\end{aligned}
$$
```{r}
to_predict$earning_wo_default <- 0.10 * to_predict$amount

```


Expected earnings : *using the law of total probability*
$$
\begin{aligned} Expected \ earning  = \;& ( 1 - p_{default} ) * Earning \ without \ default \ - \ p_{default} \ * \ Loan \ amount
\end{aligned}
$$
```{r}
to_predict$expected_earning <- (1 - to_predict$p_LPM) * to_predict$earning_wo_default - to_predict$p_LPM * to_predict$amount
```


A loan is accepted if the expected earning is positive
```{r}

# if loan accepted : Final decision = 1 
# if loan refused : Final decision = 0
to_predict$final_decision <- as.numeric(to_predict$expected_earning > 0)

print(str_glue("Number of accepted loan : {sum(to_predict$final_decision)}")) 
accepted_loan_names_list <- to_predict[to_predict$final_decision == 1,"prospect"]

print(str_glue("Accepted loan names list : {accepted_loan_names_list}"))


```


The expected total profit would be : 

```{r}
total_expected_earning = sum(as.numeric(to_predict$final_decision) * to_predict$expected_earning)

print(str_glue("expected earning : {round(total_expected_earning,2)} $"))
```

### Summary of the decision : 
```{r}
expected_profit_by_prospect <- ggplot(to_predict,
       aes(x = reorder(prospect, expected_earning),
           y = expected_earning,
           fill = factor(final_decision))) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(
    values = c("0" = "grey70", "1" = "steelblue"),
    labels = c("0" = "Rejected", "1" = "Accepted")
  ) +
  labs(
    title = "Expected Profit by Loan Applicant",
    x = "Prospect",
    y = "Expected profit ($)",
    fill = "Decision"
  ) +
  theme_minimal()

print(expected_profit_by_prospect)

ggsave( filename = "Output_plots/expected_profit_by_prospect.png", plot = expected_profit_by_prospect , width = 9, height = 6, dpi = 300 )

```


### Bad Scenario

To limit the risk at the maximum we can use the upper bound of the probability interval 

```{r}
#predict probability with prediction interval
prediction_interval <- predict(LPM,newdata = to_predict,interval="prediction",level = 0.95
)
to_predict$Bad_Scn_p_LPM <- pmin(1,pmax(0, prediction_interval[, "upr"]))

#compute expected earning 
to_predict$Bad_Scn_expected_earning <- (1 - to_predict$Bad_Scn_p_LPM) * to_predict$earning_wo_default - to_predict$Bad_Scn_p_LPM * to_predict$amount

#compute final decision
to_predict$Bad_Scn_final_decision <- as.numeric(to_predict$Bad_Scn_expected_earning > 0)

print(str_glue("Number of accepted loan : {sum(to_predict$Bad_Scn_final_decision)}")) 
Bad_Scn_accepted_loan_names_list <- to_predict[to_predict$Bad_Scn_final_decision == 1,"prospect"]

#compute total expected earning
Bad_Scn_total_expected_earning = sum(as.numeric(to_predict$Bad_Scn_final_decision) * to_predict$Bad_Scn_expected_earning)

print(str_glue("Accepted loan names list : {Bad_Scn_accepted_loan_names_list}"))

print(str_glue("expected earning : {round(Bad_Scn_total_expected_earning,2)} $"))

```

### Visual comparison between bad scenario and baseline 

```{r}
scenario_results <- data.frame(
  Scenario = c("Baseline", "Negative scenario"),
  Expected_Profit = c(
    total_expected_earning,
    Bad_Scn_total_expected_earning
  )
)

scenarii_comparison <- ggplot(scenario_results,
       aes(x = Scenario, y = Expected_Profit, fill = Scenario)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = paste0(round(Expected_Profit, 2), " $")),
    vjust = -0.5
  ) +
  labs(
    title = "Expected Portfolio Profit: Baseline vs Downside",
    x = NULL,
    y = "Expected profit ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(scenarii_comparison)

ggsave( filename = "Output_plots/scenarii_comparison.png", plot = scenarii_comparison , width = 9, height = 6, dpi = 300 )
```




